{\scshape Denicek}: Computational Substrate for Concrete, \\Collaborative, Interactive Programming

### 2024-11-11 paper.tex
Evaluation, coevolution, provenance figure

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
\vspace{-1em}
\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
the ``Remove completed'' button (2) modifies target and specifies a condition. }
\label{fig:interactive}
\vspace{-0.5em}
\end{figure}

@@

Specifying such generalization manually is cumbersome, but a programming by demonstration system
that implements typical heuristic for generalization from examples \cite{myers-2000-intelligence}
would be able to infer and suggest such generalization to the user (possibly also automatically
adding the formula based on positive and negative examples of selected items).

\begin{figure}[t]
\includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
\caption{Increment wraps the existing count in a formula that adds 1 to the
previous value (1), (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).}
\label{fig:counter}
\vspace{-0.5em}
\end{figure}

@@

@@

xx

@@

\begin{figure}[t]
  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
  \vspace{-1em}
  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
  is added (3), only the results of affected formulas are invalidated.}
  \label{fig:incremental}
  \vspace{-0.5em}
\end{figure}

\subsection{Schema code co-evolution}
\label{sec:impl-schema}

>>>>

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
\vspace{-1.25em}
\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
the ``Remove completed'' button (2) modifies target and specifies a condition. }
\label{fig:interactive}
\vspace{-2em}
\end{figure}

@@

\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
Programming by demonstration systems typically implement heuristic for generalization
\cite{myers-2000-intelligence} infers and suggests such generalizations. When integrated into
a system based on Denicek, such heuristic may also be able to automatically add a formula
based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected items).}

\begin{figure}[t]
\includegraphics[width=0.95\columnwidth,clip,trim=0cm 7cm 8cm 0cm]{fig/counter.pdf}
\caption{Increment wraps the existing count in a formula that adds 1 to the
previous value (1), (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).}
\label{fig:counter}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[t]
\centering
\vspace{-0.5em}
\begin{minipage}{0.6\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.4\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. When they conflict with
    ordinary edits (4), evaluated edits are removed (A).}
    \label{fig:eval}
  \end{minipage}
  \vspace{-1em}
\end{figure}

@@

As illustrated in Fig.~\ref{fig:incremental}, Denicek supports incremental re-computa\-tion.
In the example, the cost of speaker travel depends on the number of speakers, but the cost
of refreshments depends only on two constants. Edit that adds a speaker only invalidates the
former.

The evaluation mechanism is illustrated in Fig.~\ref{fig:eval}. When the formulas are evaluated,
Denicek generates \emph{evaluated edits} that are appended to the top of the history.
When subsequent edits are made, they are appended after all non-evaluated edits and the
evaluated edits are pushed through the newly added edits. If the edits conflict (according to
Denicek's conflict detection), affected evaluated edits and edits that depend on them are dropped.

\keyideabox{\faRefresh}{Live Programming}{In the exploratory prototype, Denicek does not
automatically evaluate formulas. This makes it easier to see and understand the evaluation
mechanism, but a realistic system based on the substrate would likely automatically evaluate
formulas to provide a live programming experience \cite{petricek-2020-foundations,rein-2019-live}.}

\begin{figure}[t]
  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
  \vspace{-0.5em}
  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
  is added (3), only the results of affected formulas are invalidated.}
  \label{fig:incremental}
\end{figure}

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
\label{fig:coevolution}
\end{figure}

\subsection{Schema Code Co-evolution}
\label{sec:impl-schema}

The problem of schema evolution is that, when the schema used in a database or a program evolves,
data and code that depend on the schema need to evolve correspondingly. The problem is well-known
in database systems \cite{rahm-2006-schema} and has recently been studied in the context of
programming systems \cite{edwards-2025-schema}.

Although Denicek does not explicitly track document structure (schema or type), all documents
have an implicit structure and some of the edits transform the document structure. As
discussed in \S\ref{sec:system-ops}, Denicek automatically updates reference nodes in the
document in this case. This enables a form of schema code co-evolution \cite{edwards-2025-schema}.
The formulas embedded in Denicek documents use reference nodes to refer to both data sources
(in the document) and the results of other computations. Consequently, if the document structure
changes, the formulas are automatically updated.

Consider the example in Fig.~\ref{fig:coevolution}. The original list {\small\Verb_<ul>_} is
turned into {\small\Verb_<tbody>_} using \ident{UpdateTag} and wrapped inside {\small\Verb_<table>_}
with a field {\small\Verb_body_} using \ident{WrapRecord}. For the latter, Denicek updates the
reference accordingly turning the original {\small\Verb_/speaker_} reference in the formula
into {\small\Verb_/speaker/body_}.

As noted earier, evaluated edits transform the formula structure (wrapping it in the
{\small\Verb_<x-evaluated>_} record). However, they are marked as non-structural and so
references to formulas are not transformed when the formula is evaluated (otherwise, references
to evaluated formulas would not point to the evaluated result, but to the original unevaluated
definition).

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\vspace{-1em}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\vspace{-0.5em}
\end{figure}

@@
